Changes in Groupings of Geometric Signs on Paleolithic Mobile Objects as a Model of Social Dynamics Across Aurignacian Time Periods

### 2025-03-05 endofquarterpaper.qmd
intro edits

# Introduction and Background

The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020] coupled with biological evolution. Another distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile. Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence, @Sauvet_2018, @Kuhn_Stiner_2007, @Vanhaeren_d]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020]. In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe across the phases of the Aurignacian period.

Social organization dynamics have previously been explored using cluster analysis methods upon variations in artefacts to represent and understand patterns of cultural groups. For example, multivariate analyses of variations in beads from the Gravettian era suggests an east to west cline of nine cultural groups [@baker2024evidence]. Seriation analysis, PCoA, and network analysis were used to identify the groups from the bead data, which consisted of 134 discrete bead variations from both burial and occupation sites, as distinguished by shape and material. The geographically discrete groups identified encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and Western Mediterranean regions. The isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance, was tested using a Mantel test, finding that geographic distance cannot solely account for the bead distribution, proving the influence of cultural and social dynamics on the groups found.

Similar analyses were performed upon personal ornament types from the Aurignacian technocomplex, with the variation in types serving as a proxy for linguistic diversity [@Vanhaeren_d]. Seriation, correspondence, and GIS analyses were used to identify groups in a personal ornament data set consisting of 157 distinct ornament types from Eastern Aurignacian sites. The analyses identified fourteen geographically cohesive groups sharing similar ornament type associations, which sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region. The possibility that the observed groups were influenced by availability of raw materials was considered, and ultimately discarded as an explanation for the patterns.

Additionally, @Sauvet_2018 performed quadrilateral geometric parietal signs to identify Upper Paleolithic cultural groups in the Cantabrian region of Spain. Factorial correspondence analysis and ascending hierarchical clustering based on 45 identified morphological characteristics were used to divide the signs into groups. Two clusters were identified from those methods, representing two distinct cultural groups. @Sauvet_2018 makes a case for the relevance of geometric signs as identity markers in identifying cultural groups, pointing to the relevant qualities of signs, as well as historical and ethnographic case studies of signs serving as markers of group membership case studies.

Following the work done by @baker2024evidence, @Vanhaeren_d, and @Sauvet_2018, in using variation in ornamentation and signs to identify cultural groups, we look to see how geometric sign design changes across the phases of the Aurignacian. The Aurignacian period occurred roughly 43,000 - 30,000 years BP, and is largely characterized by material culture indicators, as well as an ongoing biological shift from Neanderthals to early modern humans. Chronology of the Aurignacian is primarily determined by correspondence with those material culture markers, which include carinated scrapers, small-flake tool and split-based bone and ivory tools [@Chu_Richter_2020]. The Aurignacian can be divided into phases, again mainly distinguished by shifts in material culture; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015], as can be seen in @fig-aur-table. @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main phases, Aur-P1 (Proto/Early), whose typological observations corresponded with the time period of 43–37 kyr BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods kyr BP and 37-32 kyr BP. @Banks_dErrico_Zilhao_2013 used Bayesian modeling methods to further divide Aur-P1 into the Proto and Early Aurignacian, placing the Proto-Aurignacian at 41.5 - 39.9 k cal BP, and the early Aurignacian at 39.8 - 37.8 k cal BP. They found that the Proto-Aurignacian occurs during the Greenland Interstadials 9 and 10, periods of climatic amelioration, while the Early Aurignacian began with the Henrich Stadial, which is characterized by dry and arid conditions, and ended with Greenland Interstadial 8.

@@

In this study, we aim to address the following questions: How does sign distribution change across phases of the Aurignacian? How do the groupings in sign distribution vary in diversity, size, connectivity, and boundaries? What do the patterns tell us about dynamics of social organization throughout the Aurignacian? First, we identify clusters within each time period using seriation and network analysis. We then test the strength of the clusters using modularity, perMANOVA, and PERMDISP2 tests. Lastly, we use mantel tests to evaluate the null hypothesis of isolation-by-distance for each time period.

>>>>

# Introduction

The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020] coupled with biological evolution. Another distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile. Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence, @Sauvet_2018, @Kuhn_Stiner_2007, @Vanhaeren_d]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43 - 30 k BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020]. In this paper, we investugate the social organization dynamics by analyzing the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe across four phases of the Aurignacian period.

Social organization dynamics of the Upper Paleolithic have previously been explored using statistical analyses of artifacts to understand patterns of cultural groups. Multivariate analyses of beads from the Gravettian technocomplex suggests an east to west cline of nine cultural groups [@baker2024evidence]. Seriation analysis, PCoA, and network analysis were used to identify the groups from 134 discrete bead types recovered from both burial and occupation sites. To validate their claims of cultural groups, @baker2024evidence used a Mantel test to evaluate an isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance, finding that geographic distance alone did not solely account for the bead distribution. The nine geographically discrete groups encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and western Mediterranean regions.

Similar analyses of personal ornament types from the Aurignacian technocomplex identified fourteen geographically cohesive groups [@Vanhaeren_d]. Drawing on ethnographic studies that show how body decoration and ornamentation indicates ethno-linguistic identity, @Vanhaeren_d used Aurignacian personal ornamentation as a proxy for ethno-linguistic diversity. They analyzed 157 distinct ornament types from 98 Aurignacian sites in Europe and the Near East. Geographically cohesive groups sharing similar ornament type associations were identified by seriation, correspondence, and GIS analyses. These groups sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region. @Vanhaeren_d interpret the variation in personal ornament types as evidence of long-lasting cultural differences and as robustly establishing the ethno-linguistic diversity of the Aurignacian.

Cultural difference has also been identified from geometric signs in Upper Paleolithic parietal art from the Cantabrian region of Spain [@Sauvet_2018]. Forty-two quadrilateral geometric signs from four cave sites, likely associated with the Magdalenian period, spanning a distance of 30 kilometers were recorded. Multivariate analysis of 45 morphological characteristics of the signs revealed two clusters. To interpret these clusters, @Sauvet_2018 draw on case studies of geometric signs as identity markers, including South African engraved ostrich shells from 60 k BP, maker's marks on Gallo-Roman pottery, and 19th century French craft guild members' personal marks. Using these case studies as analogies, @Sauvet_2018 interpret the Magdalenian geometric signs as identity markers, and the two clusters as representative of cultural groups.

In this study, we extend the use of geometric signs as identity markers to investigate cultural groups in the Aurignacian period. The Aurignacian period spans roughly 43 - 30 k BP, and is largely characterized by material culture indicators, such as carinated scrapers, small-flake tools and split-based bone and ivory tools, as well as an ongoing evolutionary shift from Neanderthals to early modern humans [@Chu_Richter_2020]. The Aurignacian can be divided into phases, again mainly distinguished by shifts in material culture; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015], as can be seen in @fig-aur-table. @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main phases, Aur-P1 (Proto/Early), which corresponds with the time period of 43–37 k BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods 37-32 k BP. @Banks_dErrico_Zilhao_2013 used Bayesian modeling methods to further divide Aur-P1 into the Proto and Early Aurignacian, placing the Proto-Aurignacian at 41.5 - 39.9 k BP, and the early Aurignacian at 39.8 - 37.8 k BP. They found that the Proto-Aurignacian occurs during the Greenland Interstadials (GI) 9 and 10, periods of climatic amelioration, while the Early Aurignacian began with the Henrich Stadial, which is characterized by dry and arid conditions, and ends with GI 8.

@@

We hypothesize that these periods, and the environmental changes across the Aurignacian, may correspond to changes in social dynamics of cultural groups, as represented by patterns in the distribution of geometric signs. We aim to address the following questions: How do grouping dynamics of geometric signs change across the four phases of the Aurignacian? How do the groupings vary in diversity, size, connectivity, and boundaries? Lastly, what do the patterns tell us about dynamics of social organization throughout the Aurignacian? We answer these questions by first identifying clusters within each time period using seriation and network analyses. We then test the strength of the clusters using modularity, perMANOVA, and PERMDISP2 tests. Lastly, we use Mantel tests to evaluate the null hypothesis of isolation-by-distance for each time period.
